Sample-Efficient Reinforcement Learning \\ with Implicitly Quantized Representations

### 2024-02-02 paper/main.tex
updates

\section{Related Work}
\label{sec:related_work}

Model-based RL (MBRL) can significantly improve the sample efficiency of the model-free methods by reasoning about the world and maximizing the fraction of the information stream utilized for learning by utilizing the observations instead of simply trying to learn a policy or a value function to maximize the return \cite{haRecurrentWorldModels2018}. The world model can be used for planning \cite{allen1983planning, basye1992decision}. A prominent idea has been to optimize the evidence lower bound of observation and reward sequences to learn world models that operate on the latent space of a learned Variational Autoencoder  \citep[VAE,][]{kingmaAutoEncoding2014, igl2018deep}. These models rely on maximizing the conditional observation likelihood $p(o_t | s_t$), which is the reconstruction objective. The latent space of the model can be used for both planning with model-predictive control \cite{rubinstein1997optimization, hafnerLearning2019} or policy learning in the imagination of the world model \cite{hafner2019dream}. \citet{finn2017deep} and \citet{ebert2018visual} show that it is possible to plan with models that have been taught to predict videos directly without a VAE-based latent space.
% Incorporating the model uncertainty into planning through ensembling or sampling can also have a significant positive impact on the performance of model-based RL agents \cite{deisenrothPILCO2011, chua2018deep, janner2019trust, yu2020mopo, clavera2020model}.

@@

How can high-dimensional and noisy observations be used efficiently in model-based RL without resorting to reconstruction or direct prediction of future frames? A common solution has been to attach an auxiliary loss to the RL objective and perform representation learning \cite{tomar2021learning, ni2024bridging}. A common approach for learning suitable representations is the use of self-predictive abstractions, where the model is trained to predict future latent states through an auxiliary loss \cite{subramanian2022approximate}. \citet{ye2021mastering} introduce a self-supervised consistency loss on the learned latent representations. Instead of relying on a reconstruction-based loss function, \citet{schwarzer2020data} propose a cosine similarity loss between predicted future latents and the true future latents and Q-learning in the learned latent space. \citet{ma2021contrastive} modify the classical world model learning objective by replacing the observation reconstruction loss with a contrastive learning term, which maximizes the mutual information between latent states and observations and helps combat the model focusing on learning task-irrelevant information. \citet{kipf2019contrastive} use contrastive learning to learn a structured world model. \citet{ghugare2022simplifying} argue that aligning the auxiliary objectives and RL objective can be difficult. The authors derive a bound directly on the overall RL objective, which is compatible with self-consistency. The contrastive objectives for model-based RL significantly improve sample efficiency, but they are not only limited to model-based RL. \citet{laskinCURLContrastiveUnsupervised2020} propose replacing the transition function with data augmentations to get a contrastive target and performing model-free learning using the representation produced by the encoder trained with the contrastive target.

TD-MPC \cite{hansenTemporalDifferenceLearning2022} use a consistency loss to learn representations for planning with Model Predictive Path Integral control together with reward and value functions learned through temporal difference methods \cite{williams2015model}. TD-MPC2 \cite{hansen2023td} presents a series of improvements to TD-MPC to make the world model show scaling benefits and achieve state-of-the-art performance in a multi-task setting. They train world models with up to 317 million parameters and show monotone improvement in the normalized score across multiple environments and model sizes. The TD-MPC family of methods is applicable to real-world robotics \cite{lancaster2023modem} and learn representations that are suitable for transfer learning after adaptation \cite{yang2023movie} or fine-tuning \cite{feng2023finetuning}. \citet{zhu2023repo} demonstrate that the observation encoder can be made more resilient to spurious correlations by restricting the information flow from the observations to the latent representation, and the encoder can be easily aligned at test time in a reward-free manner. The loss function of TD-MPC can also be used to learn a manager in a hierarchical RL setting \cite{chitnis2023iql}.

\citet{zhaoSimplifiedTemporalConsistency2023} show that the planning component of TD-MPC is not strictly necessary for high performance, and applying model-free learning on top of the self-consistent representations is sufficient for performance competitive with state-of-the-art. \citet{yuan2022euclid} propose using intrinsic rewards for pre-training temporally consistent models before using the rewards for adapting the world model for downstream tasks. We build on top of TCRL to show that we can combat the representation collapse through a normalization term and learn task-agnostic representations by dropping the reward term of the TCRL loss and only utilizing a multistep temporal consistency loss.

@@

\textbf{Reward prediction is not necessary for representation learning}
Unlike prior methods such as TCLR \cite{zhaoSimplifiedTemporalConsistency2023}, TD-MPC \cite{hansenTemporalDifferenceLearning2022}, and Dreamer-V2 \cite{hafnerMasteringAtariDiscrete2022}, our representation learningloss (\cref{eq:rep-loss}) does not include a term for learning to predict the reward or the value. Instead, we rely solely on the self-predictive temporal consistency loss. To analyze the impact of not including the reward prediction term, we perform an ablation study, where we compare our method to three variants of our method:
\begin{itemize}
	\item \ourrew, where we add a reward prediction head into our model architecture by re-defining the dynamics as $\hat{z}_{t+1}, \hat{r}_t = d_{\phi} (z_{t}, a_{t})$ (see \cref{eq:transition}), and include a reward prediction term (discounted MSE Loss) in the representation loss:
	\begin{align*}
		\mathcal{L}_\text{rew} = \sum_{h=0}^{H-1} \gamma^h \| \hat{r}_{t+h} - r_{t+h}\|_2^2,
	\end{align*}
	where $r_{t+h}$ is the ground-truth H-step reward and $\hat{r}_{t+h}$ the predicted H-step rewards.
	\item \ourcosrew, where in addition to the reward prediction head in \ourrew, we additionally convert the self-supervised MSE loss (see \cref{eq:rep-loss}) for the temporal consistency into a cosine similarity, similarly to TCRL:
	\begin{align*} \label{eq:rep-loss}
		\mathcal{L}_{\text{rep}}(\theta, \phi; \tau)
		&= - \sum_{h=0}^{H-1} \gamma^{h} \left ( \frac{\hat{z}_{t+h}}{\| \hat{z}_{t+h} \|_2} \right )^T \left ( \frac{\tilde{z}_{t+h}}{\| \tilde{z}_{t+h} \|_2} \right ).
	\end{align*}
  % where as before we have $\hat{z}_{t+h} = d_{\phi}(\hat{z}_{t}, a_{t})$ and $\tilde{z}_{t+h} = e_{\bar{\theta}}(s_{t+1})$.
	\item TCRL-ours, where in addition to \ourcosrew, we remove the FSQ normalization scheme to make the method equal to TCRL with some minor improvements to the method, such as using nstep-TD3 \cite{fujimotoAddressingFunctionApproximation2018} as the base algorithm instead of DDPG with Double Q and nstep returns, using the neural network architecture from \citet{hansen2023td}, and performing policy smoothing \cite{fujimotoAddressingFunctionApproximation2018}.
\end{itemize}
The results for this ablation study are shown in \cref{fig:reward-ablation}. 
The plots show that our method, \our, without a reward prediction term in the loss, has equal or superior performance to the comparison methods except in Fish Swim, where the unnormalized TCRL-ours outperforms all the methods with normalization due to being able to maintain the latent matrix rank in this specific environment (see \cref{fig:rank-comparison-z1}). Even then, \our performs equally to the normalized variants of our method with the reward prediction. Our results imply that learning to predict the reward is not necessary for learning a suitable latent representation. The upside of not including the reward prediction head is that it makes the representation task-agnostic, which we believe to be important for downstream applications such as learning in an incremental multi-task setting.

@@

To provide a thorough analysis of \our, we include results where we add a reconstruction term to our representation loss \cref{eq:rep-loss}:
\begin{align*}
	\mathcal{L}_o &= \mathbb{E}_{o_t\sim\mathcal{D}} [\| \hat{o}_t - o_t \|_2^2],
\end{align*}
where $\hat{o}_t = g_\xi(z_t)$ is a learned observation decoder that takes the latent state as the input and outputs the reconstructed observation. The decoder $g_\xi$ is a standard MLP. We perform reconstruction at each time step in the horizon.
% $g_\xi$ is a learned observation decoder that takes the latent state as the input and outputs the reconstructed observation. The decoder $g_\xi$ is a standard MLP. We perform reconstruction at each time step in the horizon.
The results in \cref{fig:reconstruction-loss-ablation} show that in no environments does reconstruction aid learning, and in some environments, such as the difficult Dog Walk and Quadruped Run, including the reconstruction term has a significant detrimental effect on the performance.
This supports the observations of \citet{zhaoSimplifiedTemporalConsistency2023} and \citet{hansenTemporalDifferenceLearning2022} about the lack of need for a reconstruction target in continuous control tasks.
%Moreover, in the quadruped-run and dog-walk environments incorporating a reconstruction term has a detrimental impact on
%overall performance.

@@

\section{Conclusion}
\label{conclusion}
We have presented \our, a simple technique for learning representations which demonstrates state-of-the-art performance in
continuous control tasks.
Importantly, our normalization and bounding of the latent space empirically preserves the representation's matrix rank,
indicating that it alleviates representation collapse.
Importantly, our method is {\em (i)} straightforward, {\em (ii)} compatible with any model-free RL algorithm,
and {\em (iii)} learns a task-agnostic representation.
% Moreover, it can be combined with any model-free RL algorithm
% In our experiments we compared against both state-of-the-art model-free RL methods and state-of--the-art model-based RL methods.
% Our results demonstrate state-of-the-art based on our empirical results, we suggest further

\textbf{Limitations and future work}
To the best of our knowledge, we are the first to present a method that learns a task-agnostic representation
using only latent-state consistency. As such, we hope our method for learning task-agnostic representations paves the
way for more research into multi-task and lifelong learning.
% Given that our method obtains an implicit codebook an obvious future direction is exploring if we can leverage the
% discretized latent-space. So far, we have only applied our representation learning loss to state-based tasks. Extending the method to handle even higher-dimensional visual tasks is an interesting direction.
% Furthermore, we have neither analyzed the scaling behavior of our world models nor how well \our performs in a multi-task setting. Utilizing the knowledge learned during with Internet-scale pre-training with large robotic datasets could make our method even more sample efficient. Exploring whether explicitly quantifying the model uncertainty by \eg\ using larger ensembles would help to adapt our method to a real-world robotics setting is also a promising direction for future work.
Given that our method obtains an implicit codebook, an obvious future direction is exploring whether we can leverage the
discretized latent space. So far, we have only applied our representation loss to state-based tasks. Extending the method to handle even higher-dimensional visual tasks is an attractive direction.
% Furthermore, we have neither analyzed the scaling behavior of our world models nor how well \our performs in a multi-task setting. Utilizing the knowledge learned during with Internet-scale pre-training with large robotic datasets could make our method even more sample-efficient. Exploring whether explicitly quantifying the model uncertainty by \eg using larger ensembles would help to adapt our algorithm to a real-world robotics setting is also a promising direction for future work.

>>>>

\section{Related Work}
\label{sec:related_work}
In this section, we provide an overview of sample efficiency in RL.
Importantly, we recap how representation learning enabled model-free RL methods to obtain
state-of-the-art sample efficiency.

\textbf{Model-based RL}
Model-based RL (MBRL) is often said to be more sample efficient than model-free methods.
% This is because it can reason about the world and maximize the fraction of the information stream utilized for learning by utilizing the observations
This is because it learns a model in which it can reason about the world,
instead of simply trying to learn a policy or a value function to maximize the return \cite{haRecurrentWorldModels2018}.
% The world model can be used for planning \cite{allen1983planning, basye1992decision}.
A prominent idea has been to optimize the evidence lower bound of observation and reward sequences to learn world models that operate on the latent space of a learned Variational Autoencoder  \citep[VAE,][]{kingmaAutoEncoding2014, igl2018deep}. These models rely on maximizing the conditional observation likelihood $p(o_t | s_t$), which is the reconstruction objective.
The latent space of the model can be used for both planning with model-predictive control \cite{rubinstein1997optimization, hafnerLearning2019} or policy learning in the imagination of the world model \cite{hafner2019dream}. \citet{finn2017deep} and \citet{ebert2018visual} show that it is possible to plan with models that have been taught to predict videos directly without a VAE-based latent space.
% Incorporating the model uncertainty into planning through ensembling or sampling can also have a significant positive impact on the performance of model-based RL agents \cite{deisenrothPILCO2011, chua2018deep, janner2019trust, yu2020mopo, clavera2020model}.

@@

\textbf{Self-predictive abstractions}
How can high-dimensional and noisy observations be used efficiently in model-based RL without resorting to reconstruction or direct prediction of future frames? A common solution has been to attach an auxiliary loss to the RL objective and perform representation learning \cite{tomar2021learning, ni2024bridging}. A promising approach for learning suitable representations is the use of self-predictive abstractions, where the model is trained to predict future latent states through an auxiliary loss \cite{subramanian2022approximate}. \citet{ye2021mastering} introduce a self-supervised consistency loss on the learned latent representations. Instead of relying on a reconstruction-based loss function, \citet{schwarzer2020data} propose a cosine similarity loss between predicted future latents and the true future latents and then perform Q-learning in the
learned latent space.\looseness-2
% \citet{ma2021contrastive} modify the classical world model learning objective by replacing the observation reconstruction loss with a contrastive learning term, which maximizes the mutual information between latent states and observations, helping combat the model focusing on learning task-irrelevant information. \citet{kipf2019contrastive} use contrastive learning to learn a structured world model. \citet{ghugare2022simplifying} argue that aligning the auxiliary objectives and RL objective can be difficult. The authors derive a bound directly on the overall RL objective, which is compatible with self-consistency. The contrastive objectives for model-based RL significantly improve sample efficiency, but they are not only limited to model-based RL. \citet{laskinCURLContrastiveUnsupervised2020} propose replacing the transition function with data augmentations to get a contrastive target and performing model-free learning using the representation produced by the encoder trained with the contrastive target.

\textbf{Latent-state consistency for model-based RL}
TD-MPC \cite{hansenTemporalDifferenceLearning2022} use a consistency loss to learn representations for planning with Model Predictive Path Integral control together with reward and value functions learned through temporal difference methods \cite{williams2015model}. TD-MPC2 \cite{hansen2023td} presents a series of improvements to TD-MPC to make the world model show scaling benefits and achieve state-of-the-art performance in a multi-task setting. They train world models with up to 317 million parameters and show monotone improvement in the normalized score across multiple environments and model sizes. The TD-MPC family of methods is applicable to real-world robotics \cite{lancaster2023modem} and learn representations that are suitable for transfer learning after adaptation \cite{yang2023movie} or fine-tuning \cite{feng2023finetuning}. \citet{zhu2023repo} demonstrate that the observation encoder can be made more resilient to spurious correlations by restricting the information flow from the observations to the latent representation, and the encoder can be easily aligned at test time in a reward-free manner.
% The loss function of TD-MPC can also be used to learn a manager in a hierarchical RL setting \cite{chitnis2023iql}.

\textbf{Latent-state consistency for model-free RL}
\citet{zhaoSimplifiedTemporalConsistency2023} show that the planning component of TD-MPC is not strictly necessary for high performance, and applying model-free learning on top of the self-consistent representations is sufficient for performance competitive with state-of-the-art. \citet{yuan2022euclid} propose using intrinsic rewards for pre-training temporally consistent models before using the rewards for adapting the world model for downstream tasks. We build on top of TCRL to show that we can combat the representation collapse through a normalization term and learn task-agnostic representations by dropping the reward term of the TCRL loss and only utilizing a multistep temporal consistency loss.

@@

\textbf{Reward prediction is not necessary for representation learning}
Unlike prior methods such as TCLR \cite{zhaoSimplifiedTemporalConsistency2023}, TD-MPC \cite{hansenTemporalDifferenceLearning2022}, and Dreamer-V2 \cite{hafnerMasteringAtariDiscrete2022}, our representation learningloss (\cref{eq:rep-loss}) does not include a term for learning to predict the reward or the value. Instead, we rely solely on the self-predictive temporal consistency loss. To analyze the impact of not including the reward prediction term, we perform an ablation study, where we compare our method to three variants of our method:
\begin{itemize}
	\item \ourrew, where we add a reward prediction head into our model architecture by re-defining the dynamics as $\hat{z}_{t+1}, \hat{r}_t = d_{\phi} (z_{t}, a_{t})$ (see \cref{eq:transition}), and include a reward prediction term (discounted MSE Loss) in the representation loss:
	\begin{align*}
		\mathcal{L}_\text{rew} = \sum_{h=0}^{H-1} \gamma^h \| \hat{r}_{t+h} - r_{t+h}\|_2^2,
	\end{align*}
	where $r_{t+h}$ is the ground-truth H-step reward and $\hat{r}_{t+h}$ the predicted H-step rewards.
	\item \ourcosrew, where in addition to the reward prediction head in \ourrew, we additionally convert the self-supervised MSE loss (see \cref{eq:rep-loss}) for the temporal consistency into a cosine similarity, similarly to TCRL:
	\begin{align*} \label{eq:rep-loss}
		\mathcal{L}_{\text{rep}}(\theta, \phi; \tau)
		&= - \sum_{h=0}^{H-1} \gamma^{h} \left ( \frac{\hat{z}_{t+h}}{\| \hat{z}_{t+h} \|_2} \right )^T \left ( \frac{\tilde{z}_{t+h}}{\| \tilde{z}_{t+h} \|_2} \right ),
	\end{align*}
  where as before we have $\hat{z}_{t+h} = d_{\phi}(\hat{z}_{t}, a_{t})$ and $\tilde{z}_{t+h} = e_{\bar{\theta}}(s_{t+1})$.
	\item TCRL-ours, where in addition to \ourcosrew, we remove the FSQ normalization scheme to make the method equal to TCRL with some minor improvements to the method, such as using nstep-TD3 \cite{fujimotoAddressingFunctionApproximation2018} as the base algorithm instead of DDPG with Double Q and nstep returns, using the neural network architecture from \citet{hansen2023td}, and performing policy smoothing \cite{fujimotoAddressingFunctionApproximation2018}.
\end{itemize}
The results for this ablation study are shown in \cref{fig:reward-ablation}. 
The plots show that our method, \our, without a reward prediction term in the loss, has equal or superior performance to the comparison methods except in Fish Swim, where the unnormalized TCRL-ours outperforms all the methods with normalization due to being able to maintain the latent matrix rank in this specific environment (see \cref{fig:rank-comparison-z1}). Even then, \our performs equally to the normalized variants of our method with the reward prediction. Our results imply that learning to predict the reward is not necessary for learning a suitable latent representation. The upside of not including the reward prediction head is that it makes the representation task-agnostic, which we believe to be important for downstream applications such as learning in an incremental multi-task setting.

@@

To provide a thorough analysis of \our, we include results where we add a reconstruction term to our representation loss \cref{eq:rep-loss}:
\begin{align*}
	\mathcal{L}_o &= \mathbb{E}_{o_t\sim\mathcal{D}} [\| \hat{o}_t - o_t \|_2^2], \\
\hat{o}_t &= g_\xi(z_t),
\end{align*}
where $g_\xi$ is a learned observation decoder that takes the latent state as the input and outputs the reconstructed observation. The decoder $g_\xi$ is a standard MLP. We perform reconstruction at each time step in the horizon.
% $g_\xi$ is a learned observation decoder that takes the latent state as the input and outputs the reconstructed observation. The decoder $g_\xi$ is a standard MLP. We perform reconstruction at each time step in the horizon.
The results in \cref{fig:reconstruction-loss-ablation} show that in no environments does reconstruction aid learning, and in some environments, such as the difficult Dog Walk and Quadruped Run, including the reconstruction term has a significant detrimental effect on the performance.
This supports the observations of \citet{zhaoSimplifiedTemporalConsistency2023} and \citet{hansenTemporalDifferenceLearning2022} about the lack of need for a reconstruction target in continuous control tasks.
%Moreover, in the quadruped-run and dog-walk environments incorporating a reconstruction term has a detrimental impact on
%overall performance.

@@

\section{Conclusion}
\label{conclusion}
We have presented \our, a simple technique for learning representations which demonstrates state-of-the-art performance in
continuous control tasks.
Our normalization and bounding of the latent space empirically preserves the representation's matrix rank,
indicating that it alleviates representation collapse.
As such, \our learns a task agnostic representation and our experiments demonstrate that it shows
promise for incremental multi-task learning.
Importantly, our method is {\em (i)} straightforward, {\em (ii)} compatible with any model-free RL algorithm,
and {\em (iii)} learns a task-agnostic representation.
% Moreover, it can be combined with any model-free RL algorithm
% In our experiments we compared against both state-of-the-art model-free RL methods and state-of--the-art model-based RL methods.
% Our results demonstrate state-of-the-art based on our empirical results, we suggest further

\textbf{Limitations and future work}
To the best of our knowledge, we are the first to present a method that learns a task-agnostic representation
using only latent-state consistency. As such, we hope our method for learning task-agnostic representations paves the
way for more research into multi-task and lifelong learning.
% Given that our method obtains an implicit codebook an obvious future direction is exploring if we can leverage the
% discretized latent-space. So far, we have only applied our representation learning loss to state-based tasks. Extending the method to handle even higher-dimensional visual tasks is an interesting direction.
% Furthermore, we have neither analyzed the scaling behavior of our world models nor how well \our performs in a multi-task setting. Utilizing the knowledge learned during with Internet-scale pre-training with large robotic datasets could make our method even more sample efficient. Exploring whether explicitly quantifying the model uncertainty by \eg\ using larger ensembles would help to adapt our method to a real-world robotics setting is also a promising direction for future work.
Given that our method obtains an implicit codebook, an obvious future direction is exploring whether we can leverage the
discretized latent space. So far, we have only applied our representation loss to state-based tasks. Extending the method to handle even higher-dimensional visual tasks is an attractive direction.
Furthermore, we have neither analyzed the scaling behavior of our world models nor how well \our performs in a multi-task setting.
Utilizing the knowledge learned during with Internet-scale pre-training with large robotic datasets could make our method even more sample efficient.
% Exploring whether explicitly quantifying the model uncertainty by \eg using larger ensembles would help to adapt our algorithm to a real-world robotics setting is also a promising direction for future work.
